### 2024-08-20 paper.tex
finish cleaning up normalization proof

\section{Some linear algebra facts}
In this section we prove linear algebra facts that will be useful later.
\subsection{The centralizer of a nilpotent matrix}
\begin{definition}
    A matrix \(Y\) is \emph{Toeplitz} if it is constant along bands parallel to the main diagonal.
    That is, \(\forall i,j,k. \; Y_{ij} = Y_{i + k, j + k}\).
\end{definition}
\begin{definition}
    An \(m \times n\) matrix \(Y\) is \emph{lower-left Toeplitz} if it is Toeplitz and, in addition, we have \(y_{n - i, j - 1} = 0\) whenever \(i + j \geq \min(m,n)\).
\end{definition}
That is, \(Y\) is lower-left Toeplitz if it is Toeplitz, and the only nonzero entries are those with Manhattan distance less than \(\min(m,n)\) from the lower-left corner.
In yet other words, only the leftmost \(\min(m,n)\) diagonal bands are nonzero.

@@

\subsection{A `normalization' fact about Jordan bases}
Let \(V\) be a finite-dimensional vector space, \(A : V \to V\) a nilpotent operator, and \(f : V \to \mathbb{C}\) a linear map.
% Let \((e_{ij} : i \leq m, j \leq \lambda_i)\) be a Jordan basis for \(A\).

\begin{lemma}
    There is a Jordan basis \(e_{ij}\) for \(A\) such that there is at most one \(i\) such that there exists \(j\) such that \(f(e_{ij}) \neq 0\).
\end{lemma}
\begin{proof}
    For any Jordan basis \((e_{ij})_{ij}\) of \(A\), define
    \[S((e_{ij})_{1 \leq i \leq k, 1 \leq j \leq \lambda_i}) := \sum_i \begin{cases}-1, & \forall j. \; f (e_{ij}) = 0 \\ \lambda_i - \min\{j : f (e_{ij}) \neq 0\}, & \textrm{otherwise} \end{cases}.\]
    We proceed by induction on the measure \(S\).
    That is, let \((e_{ij})_{ij}\) be a Jordan basis for \(A\).
    Our inductive hypothesis is that if there exists a Jordan basis \((e_{ij}')_{ij}\) with \(S((e_{ij}')_{ij}) < S((e_{ij})_{ij})\), then we get the desired conclusion.
    \par Now, we have two cases.
    In the first case, \((e_{ij})_{ij}\) already satisfies the desired property.
    In this case we are done.
    In the other case, there exist \(i_1, j_1, i_2, j_2\) with \(f(e_{i_1j_1}) \neq 0\) and \(f(e_{i_2j_2}) \neq 0\).
    We let \(j_1, j_2\) be minimal with this property, so that \(\forall j < j_1. \; f(e_{i_1j}) = 0\), and \(\forall j < j_2. \; f(e_{i_2j}) = 0\).
    Wlog, we assume that \(\lambda_{i_1} - j_1 \leq \lambda_{i_2} - j_2\).
    \par By our inductive hypothesis, all we need to do is find a Jordan basis \((e_{ij}')_{ij}\) with \(S((e_{ij}')_{ij}) < S((e_{ij})_{ij})\).
    This is what we do.
    Define \(e'_{ij}\) as follows.
    \begin{itemize}
        \item \(e_{i_1, \lambda_1}' := e_{i_1, \lambda_1} - \frac{f(e_{i_1, j_1})}{f(e_{i_2, j_2})}e_{i_2, j_2 + (\lambda_{i_1} - j_1)}\)
        \item For \(j < \lambda_1\), \(e_{i_1, j}' := A^{\lambda_{i_1} - j} e_{i_1, \lambda_{i_1}}'\)
        \item For \(i \neq i_1\), \(e_{ij}' := e_{ij}\).
    \end{itemize}
    Clearly this is a Jordan basis for \(A\).
    Further, I claim that \(S((e'_{ij})_{ij})  < S((e_{ij})_{ij})\).
    It suffices to show that \(\forall j \leq j_1. \; f (e_{i_1,j}) = 0\).
    We have 
    \[f (e_{i_1,j}') = f\left(A^{\lambda_{i_1} - j}\left(e_{i_1, \lambda_1} - \frac{f(e_{i_1, j_1})}{f(e_{i_2, j_2})}e_{i_2, j_2 + (\lambda_{i_1} - j_1)}\right)\right) =\]
    \[f\left(e_{i_1, j} - \frac{f(e_{i_1, j_1})}{f(e_{i_2, j_2})} e_{i_2, j_2 + (j - j_1)}\right) = f(e_{i_1, j}) - \frac{f(e_{i_1, j_1})}{f(e_{i_2, j_2})} f(e_{i_2, j_2 + (j - j_1)}).\]
    Clearly (by design), this expression is zero when \(j = j_1\).
    And for \(j < j_1\), we have \(f(e_{i_1, j}) = f(e_{i_2, j_2 + (j - j_1)}) = 0\), so it is zero then as well.
    Thus we see that the measure \(S\) of this new basis is smaller, as desired.
\end{proof}

\begin{lemma}
    Something else; CONTINUE HERE
\end{lemma}
\begin{proof}
    Thus we see that \(S_{QP} < S_P\), contradicting that \(S_P\) is minimal.
    So there must be at most one \(i\) such that there exists \(j\) such that \(f(e_{ij}') \neq 0\).
    \par If there is no such \(i\), we have found the desired basis.
    So, suppose there is such an \(i_0\).
    Let \(U = \langle e_{i_0, j}' \rangle\).
    Simply write \(e_j := e_{i_0, j}'\).
    Let \(j_0 = \min\{j : f(e_j) \neq 0\}\).
    We just have to change basis to zero out \(e_j\) for \(j \neq j_0\).
    Let \(P_1(e_{\lambda_{i_0}}) := e_{\lambda_{i_0}} - \frac{f(e_{j_0 + 1})}{f(e_{j_0})} e_{\lambda_{i_0} - 1}\), and notice that \(f \circ P_1(e_j) = 0\) for all \(j \leq j_0 + 1\) except for \(j_0\).
    Then we define \(P_2(e_{\lambda_{i_0}}) := P_1(e_{\lambda_{i_0}}) - \frac{f\circ P_1(e_{j_0 + 2})}{f\circ P_1(e_{j_0})} e_{\lambda_{i_0} - 2}\), and notice that \(f \circ P_2(e_j) = 0\) for all \(j \leq j_0 + 2\) except for \(j_0\).    
    Eventually we get \(P_{\lambda_{i_0} - j_0}\), and by applying this to the \(e_j\)'s we obtain a basis of \(U\), in which there is exactly one \(j\) with \(f \circ P_{\lambda_{i_0} - j_0}(e_j) \neq 0\).
\end{proof}

>>>>

\section{Linear Algebra Facts}
In this section we prove linear algebra facts that will be useful later.
A reader uninterested in their verification may wish to read the theorem statements and skip ahead to the next section.

\subsection{The centralizer of a nilpotent matrix}
\begin{definition}
    A matrix \(Y\) is \emph{Toeplitz} if it is constant along bands parallel to the main diagonal.
    That is, \(\forall i,j,k. \; Y_{ij} = Y_{i + k, j + k}\).
\end{definition}
\begin{definition}
    An \(m \times n\) matrix \(Y\) is \emph{lower-left Toeplitz} if it is Toeplitz and, in addition, we have \(y_{n - i, j - 1} = 0\) whenever \(i + j \geq \min(m,n)\).
\end{definition}
That is, \(Y\) is lower-left Toeplitz if it is Toeplitz, and the only nonzero entries are those with Manhattan distance less than \(\min(m,n)\) from the lower-left corner.
In yet other words, all but the leftmost (equivalently, bottommost) \(\min(m,n)\) diagonal bands are zero.

@@

\subsection{A `normalization' fact about Jordan bases}

\begin{lemma}\label{normalization_helper_one}
    For any finite-dimensional \(V\), nilpotent \(A : V \to V\), and linear \(f : V \to \mathbb{C}\), there is a Jordan basis \(e_{ij}\) for \(A\) such that there is at most one \(i\) such that there exists \(j\) such that \(f(e_{ij}) \neq 0\).
\end{lemma}
\begin{proof}
    For any Jordan basis \((e_{ij})_{ij}\) of \(A\), define
    \[S((e_{ij})_{1 \leq i \leq k, 1 \leq j \leq \lambda_i}) := \sum_i \begin{cases}-1, & \forall j. \; f (e_{ij}) = 0 \\ \lambda_i - \min\{j : f (e_{ij}) \neq 0\}, & \textrm{otherwise} \end{cases}.\]
    We proceed by induction on the measure \(S\).
    That is, let \((e_{ij})_{ij}\) be a Jordan basis for \(A\).
    Our inductive hypothesis is that if there exists a Jordan basis \((e_{ij}')_{ij}\) with \(S((e_{ij}')_{ij}) < S((e_{ij})_{ij})\), then we get the desired conclusion.
    \par Now, we have two cases.
    In the first case, \((e_{ij})_{ij}\) already satisfies the desired property.
    In this case we are done.
    In the other case, there exist \(i_1, j_1, i_2, j_2\) with \(i_1 \neq i_2\), and \(f(e_{i_1j_1}) \neq 0\), and \(f(e_{i_2j_2}) \neq 0\).
    We let \(j_1, j_2\) be minimal with this property, so that \(\forall j < j_1. \; f(e_{i_1j}) = 0\), and \(\forall j < j_2. \; f(e_{i_2j}) = 0\).
    Wlog, we assume that \(\lambda_{i_1} - j_1 \leq \lambda_{i_2} - j_2\).
    \par By our inductive hypothesis, all we need to do is find a Jordan basis \((e_{ij}')_{ij}\) with \(S((e_{ij}')_{ij}) < S((e_{ij})_{ij})\).
    This is what we do.
    Define \(e'_{ij}\) as follows.
    \begin{itemize}
        \item \(e_{i_1, \lambda_1}' := e_{i_1, \lambda_1} - \frac{f(e_{i_1, j_1})}{f(e_{i_2, j_2})}e_{i_2, j_2 + (\lambda_{i_1} - j_1)}\)
        \item For \(j < \lambda_1\), \(e_{i_1, j}' := A^{\lambda_{i_1} - j} e_{i_1, \lambda_{i_1}}'\)
        \item For \(i \neq i_1\), \(e_{ij}' := e_{ij}\).
    \end{itemize}
    Clearly this is a Jordan basis for \(A\).
    Further, I claim that \(S((e'_{ij})_{ij})  < S((e_{ij})_{ij})\).
    It suffices to show that \(\forall j \leq j_1. \; f (e_{i_1,j}) = 0\).
    We have 
    \[f (e_{i_1,j}') = f\left(A^{\lambda_{i_1} - j}\left(e_{i_1, \lambda_1} - \frac{f(e_{i_1, j_1})}{f(e_{i_2, j_2})}e_{i_2, j_2 + (\lambda_{i_1} - j_1)}\right)\right) =\]
    \[f\left(e_{i_1, j} - \frac{f(e_{i_1, j_1})}{f(e_{i_2, j_2})} e_{i_2, j_2 + (j - j_1)}\right) = f(e_{i_1, j}) - \frac{f(e_{i_1, j_1})}{f(e_{i_2, j_2})} f(e_{i_2, j_2 + (j - j_1)}).\]
    Clearly (by design), this expression is zero when \(j = j_1\).
    And for \(j < j_1\), we have \(f(e_{i_1, j}) = f(e_{i_2, j_2 + (j - j_1)}) = 0\), so it is zero then as well.
    Hence the measure \(S\) of this new basis is smaller, as desired.
\end{proof}

\begin{lemma}\label{normalization_helper_two}
    For any \(n\) and linear \(f : \mathbb{C}^n \to \mathbb{C}\), there is a Jordan basis \(e_j\) for \(J_n\) such that there is at most one \(j\) with \(f(e_j) \neq 0\).
\end{lemma}
\begin{proof}
    Let \(e_j\) be a Jordan basis for \(J_n\).
    If \(\{j : f(e_j) \neq 0\}\) is the empty set, we are done.
    Otherwise, let \(j_0 = \min\{j : f(e_j) \neq 0\}\).
    For any Jordan basis \(f_j\) with \(j_0 = \min\{j : f(e_j) \neq 0\}\), define 
    \[S((f_j)_j) := \begin{cases}
        -1, & \{j > j_0 : f(e_j) \neq 0\} = \emptyset \\
        n - \min\{j > j_0 : f(e_j) \neq 0\}, & \textrm{otherwise}
    \end{cases}.\]
    We proceed by induction on \(S\).
    That is, let \((e_j)_j\) be a Jordan basis for \(J_n\) with \(j_0 = \min\{j : f(e_j) \neq 0\}\).
    Our inductive hypothesis is that if there exists a Jordan basis \((e_j')_j\) with \(j_0 = \min\{j : f(e_j') \neq 0\}\) and \(S((e_j')_j) < S((e_j)_j)\), then the conclusion holds.
    \par We have two cases: either \((e_j)_j\) satisfies the desired property, or not.
    If not, then let \(j_1 = \min\{j > j_0 : f(e_j) \neq 0\}\), and define a new Jordan basis \(e_j'\) as follows.
    \begin{itemize}
        \item \(e_n' := e_n - \frac{f(e_{j_1})}{f(e_{j_0})} e_{n - (j_1 - j_0)}\)
        \item For \(j < n\), \(e_j' := J_n^{n - j} e_n'\)
    \end{itemize}
    It is straightforward to check that \(j_0 = \min\{j : f(e_j) \neq 0\}\), and that \(S((e_j')_j) \leq S((e_j)_j) - 1\).
    By our inductive hypothesis, we are done.
\end{proof}

\begin{theorem}\label{normalization}
    For any finite-dimensional \(V\), nilpotent \(A : V \to V\), and linear \(f : V \to \mathbb{C}\), there is a Jordan basis \(e_{ij}\) for \(A\) such that there is at most one pair \((i, j)\) with \(f(e_{ij}) \neq 0\).
\end{theorem}
\begin{proof}
    \cref{normalization_helper_one} gives us a Jordan basis \(e_{ij}\) such that for all \(i \neq i_0\) and all \(j\), we have \(f(e_{ij}) = 0\).
    Restricting \(A\) to \(\langle e_{i_0j}\rangle_{j \textrm{ arbitrary}}\) gives a Jordan block, and then applying \cref{normalization_helper_two} gives the desired result.
\end{proof}
